---
title: "PREDICTING PAROLE VIOLATORS"
output: html_document
---

PREDICTING PAROLE VIOLATORS

In many criminal justice systems around the world, inmates deemed not to be a threat to society are released from prison under the parole system prior to completing their sentence. They are still considered to be serving their sentence while on parole, and they can be returned to prison if they violate the terms of their parole.

Parole boards are charged with identifying which inmates are good candidates for release on parole. They seek to release inmates who will not commit additional crimes after release. In this problem, we will build and validate a model that predicts if an inmate will violate the terms of his or her parole. Such a model could be useful to a parole board when deciding to approve or deny an application for parole.

For this prediction task, we will use data from the United States 2004 National Corrections Reporting Program, a nationwide census of parole releases that occurred during 2004. We limited our focus to parolees who served no more than 6 months in prison and whose maximum sentence for all charges did not exceed 18 months. The dataset contains all such parolees who either successfully completed their term of parole during 2004 or those who violated the terms of their parole during that year. The dataset contains the following variables:

male: 1 if the parolee is male, 0 if female
race: 1 if the parolee is white, 2 otherwise
age: the parolee's age (in years) when he or she was released from prison
state: a code for the parolee's state. 2 is Kentucky, 3 is Louisiana, 4 is Virginia, and 1 is any other state. The three states were selected due to having a high representation in the dataset.
time.served: the number of months the parolee served in prison (limited by the inclusion criteria to not exceed 6 months).
max.sentence: the maximum sentence length for all charges, in months (limited by the inclusion criteria to not exceed 18 months).
multiple.offenses: 1 if the parolee was incarcerated for multiple offenses, 0 otherwise.
crime: a code for the parolee's main crime leading to incarceration. 2 is larceny, 3 is drug-related crime, 4 is driving-related crime, and 1 is any other crime.
violator: 1 if the parolee violated the parole, and 0 if the parolee completed the parole without violation.

```{r}
setwd("c:/Work/Analytics-Edge/Data")
```

##Loading and Preparing the Dataset
```{r}
# Read the data set
parole <- read.csv("parole.csv")

# Structure and summary
str(parole)
summary(parole)

# Number of violators
table(parole$violator)[2]


str(parole)

```


Problem 2.2 
In the last subproblem, we identified variables that are unordered factors with at least 3 levels, so we need to convert them to factors for our prediction problem (we introduced this idea in the "Reading Test Scores" problem last week). Using the as.factor() function, convert these variables to factors. Keep in mind that we are not changing the values, just the way R understands them (the values are still numbers).
```{r}
# Convert to factors
parole$state = as.factor(parole$state)
parole$crime = as.factor(parole$crime)
summary(parole$state) 
summary(parole$crime)
```


#Training and Testing Datasets
```{r}
set.seed(144)
library(caTools)
split = sample.split(parole$violator, SplitRatio = 0.7)
train = subset(parole, split == TRUE)
test = subset(parole, split == FALSE)
```

#Building the Logistic Regression Model
Using glm (and remembering the parameter family="binomial"), train a logistic regression model on the training set. Your dependent variable is "violator", and you should use all of the other variables as independent variables.

What variables are significant in this model? Significant variables should have a least one star, or should have a probability less than 0.05 (the column Pr(>|z|) in the summary output). 
```{r}
mod = glm(violator ~ ., data=train, family="binomial")
summary(mod)

# Significant predictors
id = which(summary(mod)$coeff[,4] < 0.05)
```

What can we say based on the coefficient of the multiple.offenses variable?

The following two properties might be useful to you when answering this question:

1) If we have a coefficient c for a variable, then that means the log odds (or Logit) are increased by c for a unit increase in the variable.

2) If we have a coefficient c for a variable, then that means the odds are multiplied by e^c for a unit increase in the variable.

Our model predicts that parolees who committed multiple offenses have 1.61 times higher odds of being a violator than the average parolee.       Our model predicts that a parolee who committed multiple offenses has 1.61 times higher odds of being a violator than a parolee who did not commit multiple offenses but is otherwise identical.     Our model predicts that parolees who committed multiple offenses have 5.01 times higher odds of being a violator than the average parolee.       Our model predicts that a parolee who committed multiple offenses has 5.01 times higher odds of being a violator than a parolee who did not commit multiple offenses but is otherwise identical.

```{r}
# Cofficients of the significant predictors
coeff.sig = summary(mod)$coeff[,1][id]
```

Consider a parolee who is male, of white race, aged 50 years at prison release, from the state of Maryland, served 3 months, had a maximum sentence of 12 months, did not commit multiple offenses, and committed a larceny. Answer the following questions based on the model's predictions for this individual. (HINT: You should use the coefficients of your model, the Logistic Response Function, and the Odds equation to solve this problem.)

EXPLANATION

From the logistic regression equation, we have log(odds) = -4.2411574 + 0.3869904*male + 0.8867192*race - 0.0001756*age + 0.4433007*state2 + 0.8349797*state3 - 3.3967878*state4 - 0.1238867*time.served + 0.0802954*max.sentence + 1.6119919*multiple.offenses + 0.6837143*crime2 - 0.2781054*crime3 - 0.0117627*crime4. This parolee has male=1, race=1, age=50, state2=0, state3=0, state4=0, time.served=3, max.sentence=12, multiple.offenses=0, crime2=1, crime3=0, crime4=0. We conclude that log(odds) = -1.700629.

Therefore, the odds ratio is exp(-1.700629) = 0.183, and the predicted probability of violation is 1/(1+exp(1.700629)) = 0.154.
```{r}
odds = as.numeric(exp(coefficients(mod)[c("(Intercept)")]+coefficients(mod)[c("male")]+coefficients(mod)[c("race")] + coefficients(mod)[c("time.served")]*3 + coefficients(mod)[c("max.sentence")]*12 + coefficients(mod)[c("multiple.offenses")]*0 + coefficients(mod)[c("crime2")]))
p = odds/(1+odds)
```



#Evaluating the Model on the Testing Set
```{r}
# Predict the probability of violation for the test set
testPred = predict(mod, newdata=test, type="response")

# Confusion Matrix
table(test$violator, testPred > 0.5)
# Sensitivity, Specificity and Accuracy
12/23; 167/179; 179/202

# Baseline model
table(test$violator)
# Accuracy: there are 179 negative examples, which are the ones that the baseline model would get correct.
179/202

# Load ROCR library.  The AUC deals with differentiating between a randomly selected positive and negative example. It is independent of the regression cutoff selected. 
library(ROCR)
ROCRpred = prediction(testPred, test$violator)
as.numeric(performance(ROCRpred, "auc")@y.values)
```

